Co-methylation

### 2014-10-07 Rmarkdown/comethylation.Rmd
Review co-methylation of Lister (2009).
TODOs on re-structuring of chapter.

@@

@@

@@

@@

Using bisulfite-sequencing technologies we can analyse co-methylation at the level of individual DNA molecules. I call this within-fragment co-methylation, where fragment refers to a fragment of DNA. More precisely, within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment. It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$. For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure. We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.

@@

\subsection{Other related work}
\label{sec:previous_other_related_work}

Other definitions of co-methylation and related work:

* Lister's mCG correlations
* co-methylation across samples
* "vertical" correlations of \cite{Capra:2014uh}

@@

The method I developed, which is described in Section \ref{subsec:wf_cometh}, is superior because of __TODO__.

>>>>

__TODO: Break out review of co-methylation into its own chapter.__

@@

__TODO: Incorporate material in "Correlations" section of `a_statistical_model_of_methylC-seq.md`.__

@@

The analysis in \cite{Lister:2009hy} takes a different approach. It too uses within-sample aggregate methylation levels, but rather than estimating correlations of the methylation levels, instead looks for patterns in the spacing of "methylcytosines" along the genome.

@@

#### \cite{Lister:2009hy}

\cite{Lister:2009hy} report an $8-10$ nucleotide periodicity in the occurence "methylcytosines". Recall that \cite{Lister:2009hy} define a "methylcytosine" as a cytosine that has a non-zero $\beta$-value, as determined by a Binomial test (see Section \ref{sec:criticism_of_methylcytosines} for details and a criticism of this concept). Hence, while this may appear to be a within-fragment analysis of co-methylation, it in fact uses aggregate methylation levels. However, unlike the methods described in Section \ref{sec:previous_cor_of_agg_methylation}, the analysis of these aggregate methylation levels is not based on a correlation of these levels, but rather on unnormalised counts, which leads to problems.

\cite{Lister:2009hy} tabulate the number of neighbouring[^non_redundant] pairs of "methylcytosines" as a function of the $IPD$. These counts are plotted as a bar chart, with $IPD$ along the x-axis and count on the y-axis, along with a cubic spline fit to the counts. This is done separately for each combination of methylation context (CG, CHG and CHH) and genomic context (exonic, intronic or random) for a total of nine plots (Supplementary Figure 9 \cite{Lister:2009hy}).

[^non_redundant]: I think that "neighbouring" is what is meant by "non-redundant pair-wise distances" in the caption of Supplementary Figure 9, but I am not certain.

An $8-10$ nucleotide periodicity in the fitted cubic spline is observed in most, but not all, of these nine plots. This periodicity is very weak several contexts and not at all present for exonic CpGs. The weakness of the signal is partly attributable to the small sample sizes in the CHG and CHH contexts since there are very few "methylcytosines" for these contexts in these human samples. Furthermore, in all nine contexts the reported $8-10$ nucleotide periodicity is weaker than a $3$ nucleotide periodicity for "methylcytosines" at exonic CpGs; this $3$ nucleotide periodicity can clearly be observed in the raw counts, unlike the $8-10$ nucleotide periodicity, but is not commented on by the authors.

Based on these nine plots, \cite{Lister:2009hy} draw parallels to the $10$ periodicity in the autocorrelation of CHH $\beta$-values observed by \cite{Cokus:2008fc} and speculate that "_DNMT3A_  may be responsible for catalysing the methylation at non-CG sites". However, I suspect that the results in these nine plots are driven by periodicities in where cytosines are located in the genome rather than any periodicities in the actual methylation of these cytosines. That is not to say that there aren't periodicities in the methylation of cytosines, but that is not what is being measured in this analysis. To explain, these plots use the raw counts of "methylcytosines", which are not adjusted for the frequency of pairs of cytosines with a given $IPD$. If pairs of cytosines are observed more often at certain $IPD$s then even if methylation was totally independent of $IPD$ we would expect more pairs of "methylcytosines" at these same $IPD$s. Therefore, at a minimum, it is necessary to normalise the number of "methylcytosines" by the number of cytosines.

In fact, there are well known periodicities in the distribution of nucleotides in genomes. For example, in eukaryotic genomes there is a $10$ periodicity of G+C dinucleotides that acts out of phase with a $10$ nucleotide periodicity of A+T dinucleotides for nucleosome-bound DNA \cite{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \cite{Li:1997ts}; the former likely explains the $8-10$ nucleotide periodicity and the latter the $3$ nucleotide periodicity for exonic CpGs.

#### \cite{Landan:2012kp}

__TODO: Review co-methylation work in this paper__

@@

Using bisulfite-sequencing technologies we can analyse co-methylation at the level of individual DNA molecules. I call this within-fragment co-methylation, where a "fragment" refers to a fragment of DNA. More precisely, within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment. It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$. For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure. We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.

@@

\subsection{Between-sample co-methylation}
\label{sec:between_sample_cometh}

As is clear from the above, there are multiple definitions of within-sample co-methylation. An fundamentally different concept to within-sample co-methylation is between-sample co-methylation, which is the correlation of methylation levels between a single pair of loci across a set of samples rather than the correlation of pairs of loci within a single sample. (__CITE__) estimate various forms of between-sample co-methylation and are reviewed below.

* "vertical" correlations of \cite{Capra:2014uh}
* MethLD of \cite{Liu:dy}

__TODO: Comment on utility of within-sample vs. between-sample co-methylation__

@@

\subsection{Exploiting co-methylation in analyses of differential methylation}
\label{sec:exploiting_cometh}

\cite{Hansen:2011gu, Hebestreit:2013ko, Wang:2011cw, Pedersen:2012vl, Sofer:2013bk} have each sought to exploit co-methylation, both the within-sample and between-sample forms, in developing more powerful and efficient ways to identify differential methylation.
